SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-11 Manuscript/manuscriptPSsuppl.tex
Section S2

\section{Electrometer concept in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of negatively charged PS and PG lipids
are also systematic, but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
it is important to note that the order parameters
increase due to the addition of negative charged lipids according to the electrometer
concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
than in pure lipid bilayers. This is evident also in the headgroup order parameters
of POPC in bilayers with different amounts anionic lipids without added calcium (Fig. \ref{OrderParametersWithCaCl}).
Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50--300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
the negative charge lipids, resulting to a neutral membrane. 
Above such concentrations, the specific binding of calcium leads to
the overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl$_2$ concentration from experiments 
    with different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1~M.
    Data for Pure DPPC from Ref.~\citenum{akutsu81},
    for pure POPC from Ref.~\citenum{altenbach84}, 
    for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1~M NaCl from Ref.~\citenum{macdonald87}
    and for POPC:POPG (1:1) mixture data without NaCl from Ref.\citenum{borle85}.
  }
\end{figure}

Because the POPC headgroup order parameters in mixtures with different amounts of anionic lipids
but without additional salt are not equal, the binding affinity of calcium can be better compared
by plotting the changes of order parameters as a function of added calcium.
As expected, such a plot reveals more pronounced order parameter decrease in systems
with larger fractions of negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}),
indicating an increase in the calcium binding affinity with the increasing amount of negatively charged
lipids in membranes. In conclusion, the presented empirical comparison of headgroup order parameter changes
from various mixtures of POPC and anionic lipids with added calcium gives physically
consistent results, suggesting that the electrometer concept can be used to determine
the cation binding affinity also to the lipid bilayers containing mixtures of PC and anionic lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes of POPC headgroup order parameters as a function of CaCl$_2$
    measured from mixed bilayers containing different amounts of anionic lipids.
    The original data is the same as in figure~\ref{OrderParametersWithCaCl}.
%    The values are taken from 2H NMR experiments reported in the
%    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
%    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
%    and POPC:POPG (1:1) without NaCl \cite{borle85}).
  }
\end{figure}

>>>>

\section{Molecular electrometer in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}

The molecular electrometer concept is based on the empirical observations that
the C--H bond order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of negatively charged PS and PG lipids
are also systematic, but as they are less well characterized~\cite{borle85,macdonald87,roux86,roux90} than for PC,
the ion binding affinity to negatively charged bilayers
can be better characterized by measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters $S^\beta_\mathrm{CH}$ and $S^\alpha_\mathrm{CH}$ of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol, and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
it is important to note that these $S_\mathrm{CH}$
increase due to the addition of negative charged lipids according to the molecular electrometer~\cite{seelig87, scherer87}
(Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
than in pure PC lipid bilayers. This is demonstrated in Fig. \ref{OrderParametersWithCaCl} by the headgroup $S_\mathrm{CH}$
of POPC in bilayers with different amounts of anionic lipids without added calcium.
Upon addition of CaCl$_2$, the $S_\mathrm{CH}$ decrease and reach the values of pure PC bilayers around 
CaCl$_2$ concentrations of $\sim$50--300~mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of the bound Ca$^{2+}$ cancels
the negative charge of the lipids, resulting in a neutral membrane. 
Above these concentrations, the specific binding of calcium leads to
overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl$_2$ concentration from experiments 
    with different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1~M.
    Data for pure DPPC from Ref.~\citenum{akutsu81},
    for pure POPC from Ref.~\citenum{altenbach84}, 
    for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1~M NaCl from Ref.~\citenum{macdonald87},
    and for POPC:POPG (1:1) mixture without NaCl from Ref.\citenum{borle85}.
  }
\end{figure}

Because the POPC headgroup $S_\mathrm{CH}$ in mixtures with different amounts of anionic lipids
but without additional salt are not equal, the binding affinity of calcium can be better compared
by plotting the changes of order parameters, $\Delta S_\mathrm{CH}$, as a function of added calcium.
As expected, such a plot reveals more pronounced order parameter decrease in systems
with larger fractions of negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}),
indicating an increase in the calcium binding affinity with the increasing amount of negatively charged
lipids in membranes. In conclusion, the presented empirical comparison of headgroup order parameter changes
from various mixtures of POPC and anionic lipids with added calcium gives physically
consistent results, suggesting that the molecular electrometer can be used to determine
the cation binding affinity also to lipid bilayers containing mixtures of PC and anionic lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes of POPC headgroup order parameters, $\Delta S^\beta_\mathrm{CH}$ and $\Delta S^\alpha_\mathrm{CH}$, as a function of CaCl$_2$
    measured from mixed bilayers containing different amounts of anionic lipids.
    The original data are the same as in Fig.~\ref{OrderParametersWithCaCl}.
%    The values are taken from 2H NMR experiments reported in the
%    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
%    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
%    and POPC:POPG (1:1) without NaCl \cite{borle85}).
  }
\end{figure}
